\Title

### 2013-01-30 synmut.tex
corrected time-dep sims and figure adaptations
rewrote discussion and abstract

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp, twocolumn]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
\usepackage{color}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{natbib}
\usepackage{pslatex}
\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\pfix}{P_{fix}}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\env}{\textit{env}}
\newcommand{\shankaregion}{C2-C5}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\title{\Title}
\author{\Author}
\date{\today}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract}
\noindent
Intrapatient HIV evolution is governed by selection on the protein level in the
arms race with the immune system (killer T-cells and antibodies). Synonymous
mutations do not have an immunity-related phenotype and are often assumed to be
neutral. In this paper, we show that synonymous changes in epitope-rich regions
are often deleterious but still reach high frequencies in the viral population.  We analyze time
series of viral sequences from the \shankaregion~part of {\it env} within individual
hosts and observe that synonymous derived alleles rarely reach fixation.
Simulations suggest that such synonymous mutations
have a (Malthusian) selection coefficient of the order of $-0.001$, and that
they are brought up to high frequency by hitchhiking on neighboring beneficial
nonsynonymous alleles. We detect a negative correlation between the fixation of an allele and
its involvement in RNA stem-loop structures.
Deleterious synonymous mutations are not observed as abundantly in other parts of the HIV genome, in which
selective sweeps are less dense and hitchhiking not as strong; this behaviour is
confirmed by extensive computer simulations.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{Shankarappa_fix_loss_dt_times}
\label{fig:fixp1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{Bunnik2008_fixmid_syn_ShankanonShanka}
\label{fig:fixp2}}
\caption{Fixation and loss of synonymous mutations. Panel A shows the time
course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$. 
In each frequency interval, the  fraction of synonymous
mutations that ultimately fix is the fixation probability conditional on the
initial frequency.
Panel B) Fixation probability of derived synonymous
alleles as a function of $\nu_0$ in C2-V5 is below the neutral
expectation indicated by the diagonal line. This suppression is not
oberserved in other parts of {\it env} or for nonsynonymous mutations.
The horizontal error bars on the abscissa are bin sizes, the vertical ones the
standard deviation after 100 patient bootstraps of the data. Data from
Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.48\linewidth]{reactivities_histograms_syn}
\label{fig:SHAPEA}}
\subfloat{\includegraphics[width=0.48\linewidth]{Shankarappa_fixmid_syn_V_regions.pdf}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations have
high SHAPE reactivities.
Panel A) Synonymous substitions have
substantially higher SHAPE reactivities than mutations that reach frequencies above 15\%, but are
susequently lost (restricted to the vicinity of variable loops). This difference
is quantified by a cumulative histogram ($p=0.002$, Kolmogorov-Smirnov two 
sample test). Mutations that are never
observered show an intermediate distribution of SHAPE values.
Panel B shows the fixation probability of synonymous mutations in C2-V5
separately for variable loops and the connecting conserved regions, that
harbor conserved RNA stems. As expected, fixation probability is lower inside
the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

@@

\section{Discussion}
Despite several known functional roles for RNA secondary structure in the HIV
genome, synonymous mutations are often used as approximately neutral markers in
evolutionary studies of viruses. We have shown that the majority of synonymous
mutations in the conserved regions C2-C5 of the \env~gene are deleterious.
Comparison with recent biochemical studies of binding propensity of bases in RNA
genome suggest that these mutations are deleterious, at least in part, because they disrupt
stems in RNA secondary structures. Furthermore, we provide evidence that these
mutations are brought to high frequency through linkage to adaptive mutations.
The latter mutations are only transiently adaptive, either through a
coevolution with the immune system or redundant escape within an epitope.

Our observations and conclusion rely heavily on longitudinal data in which the
dynamics of mutations can be explicitly observed. The fact that deleterious
mutations can be brought to high frequencies through hitchhiking underscores
the intensity of the coevolution with the immune system. The fact that
multiple escape mutations in the same epitope -- as is indeed observed in
studies of antibody escape~\citep{moore_limited_2009, bar_early_2012} -- are
necessary to explain the patterns of fixation of nonsynonymous mutations points
towards a large populations size that rapidly discovers adaptive mutations. A
similar point has been made recently by Boltz {\it et al.} in the context of
preexisting drug resistance mutations~\citep{boltz_ultrasensitive_2012}.

The observed hitchhiking highlights the importance of linkage due to infrequent
recombination for the evolution of HIV
\citep{neher_recombination_2010,batorsky_estimate_2011,
josefsson_majority_2011}. The recombination rate has been estimated to be on the
order of $\rho = 10^{-5}$ per base and day. It takes roughly $t_{sw} =
\epsilon^{-1} \log \nu_0$ generations for escape mutation with escape rate $\epsilon$ to rise
from an initially low frequency $\nu_0\sim \mu$ to frequency one. This implies
that a region of length $l = (\rho t_{sw})^{-1} = \epsilon / \rho \log \nu_0$
remains linked to the adaptive mutation. With $\epsilon=0.01$, we have $l\approx
100$ bases. Hence we expect strong linkage between the variable loops and their
surrounding stems, but none far beyond the variable regions, consistent with the lack of signal
outside of C1-V5. In case of much stronger selection -- such as observed during
early CTL escape or drug resistance evolution -- the linked  region is of course
much larger \citep{nijhuis_stochastic_1998}.

A functional significance of the insulating RNA structure stems between the
hyper variable loops has been proposed
previously~\citep{watts_architecture_2009, sanjuan_interplay_2011}.
\citet{sanjuan_interplay_2011} have shown that insulating stems are relevant for
viral fitness {\it in vivo}. Our analysis is limited by the availability of
longitudinal data which requires a focus on the the variable regions of \env.
Conserved RNA structures exist in different parts of the HIV genome (several are
known). In absence of repeated adaptive substitutions in the vicinity that cause
hitchhiking, the deleterious synonymous mutations remain at low frequencies and
can only be observed by deep sequencing methods.

@@

Our results emphasize the inadequacy of independent site
models of HIV evolution and the common assumption that selection is time
independent or additive. If genetic variation is only transiently beneficial,
existing methods to quantify selection will yield substantial underestimates
\citep{williamson_adaptation_2003,neher_rate_2010,batorsky_estimate_2011}. To explain the
observations regarding the fixation probabilities of non-synonymous mutations,
either transient selection, or substational within-epitope competition are
necessary. Which mechanism is more widespread is not clear as of now,
there is evidence for both~\citep{richman_rapid_2003, moore_limited_2009,
bar_early_2012}.

\section{Methods}
\subsection{Sequence data collection}
Longitudinal intrapatient viral RNA sequences were collected for published
studies~\citep{shankarappa_consistent_1999,
liu_selection_2006, bunnik_autologous_2008} and downloaded from the Los Alamos
National Laboratory (LANL) HIV sequence database~\citep{LANL2012}. The sequences from
some patients showed signs of HIV compartimentalization into subpopulations and
were discarded; a grand total of 11
patients with approximately 6 time points each and 10 sequences per time point
were analyzed. The time interval or resolution between two consecutive sequences
was approximately 6 to 18 months.

@@

\subsection{Sequence analysis}
The good sequences were aligned within each patient
via the translated amino acid sequence, using
Muscle~\citep{edgar_muscle:_2004}, and to the NL4-3 reference sequence used
by \citet{watts_architecture_2009} in the SHAPE assay. Within each patient, a
consensus RNA sequence at the first time point was used to classify alleles as ancestral or
derived at all sites. Problematic sites that included large frequencies of gaps
were excluded from the analysis to avoid artefactual substitutions due to
alignment errors. Time series of allele frequencies were extracted from the
sequences.

@@

\section*{Acknowledgements}
\comment{to be written\dots}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp, twocolumn]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
\usepackage{color}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{natbib}
\usepackage{pslatex}
\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\pfix}{P_{fix}}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\env}{\textit{env}}
\newcommand{\shankaregion}{C2-V5}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\title{\Title}
\author{\Author}
\date{\today}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract}
\noindent

Intrapatient HIV evolution is dominated by selection on the protein level in the
arms race with the immune system. Synonymous mutations do not have an
immunity-related phenotype and are often assumed to be neutral. Here, we show
that synonymous changes in epitope-rich regions are often deleterious but still
reach high frequencies in the viral population.
We analyze longitudinal intra-patient data from the \shankaregion~part of the
envelope gene (\env) and observe that synonymous derived alleles rarely reach
fixation.
Simulations suggest that such synonymous mutations have a (Malthusian) selection
coefficient of the order of $-0.002$, and that they are brought up to high
frequency by hitchhiking on neighboring beneficial nonsynonymous alleles. We
detect a negative correlation between the fixation of an allele and its
involvement in RNA stem-loop structures.
The patterns of fixation of non-synonymous mutations suggest that antibody
escape mutations in \shankaregion~are only transiently beneficial, either since
the immune system is catching up or because of competition between equivalent
escapes.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{Shankarappa_fix_loss_dt_times}
\label{fig:fixp1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities}
\label{fig:fixp2}}
\caption{Fixation and loss of synonymous mutations. Panel A shows the time
course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$. 
In each frequency interval, the  fraction of synonymous
mutations that ultimately fix is the fixation probability conditional on the
initial frequency.
Panel B) Fixation probability of derived synonymous
alleles as a function of $\nu_0$ in C2-V5 is below the neutral
expectation indicated by the diagonal line. This suppression is not
oberserved in other parts of {\it env} or for nonsynonymous mutations.
The horizontal error bars on the abscissa are bin sizes, the vertical ones the
standard deviation after 100 patient bootstraps of the data. Data from
Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[height=0.46\linewidth]{reactivities_histograms_syn}
\label{fig:SHAPEA}}
\subfloat{\includegraphics[height=0.46\linewidth]{fixation_probabilities_VnonV}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations have
high SHAPE reactivities.
Panel A) Synonymous substitions have
substantially higher SHAPE reactivities than mutations that reach frequencies above 15\%, but are
susequently lost (restricted to the vicinity of variable loops). This difference
is quantified by a cumulative histogram ($p=0.002$, Kolmogorov-Smirnov two 
sample test). Mutations that are never
observered show an intermediate distribution of SHAPE values.
Panel B shows the fixation probability of synonymous mutations in C2-V5
separately for variable loops and the connecting conserved regions, that
harbor conserved RNA stems. As expected, fixation probability is lower inside
the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

@@

\section{Discussion}
Despite several known functional roles for RNA secondary structure in the HIV
genome, synonymous mutations are often used as approximately neutral markers in
evolutionary studies of viruses. By analyzing the fate of mutations in
longitudinal data of HIV \env{} evolution, we have shown that the majority of
synonymous mutations in the conserved regions C2-C5 of the \env~gene are deleterious.
Comparison with recent biochemical studies of binding propensity of bases in RNA
genome suggest that these mutations are deleterious, at least in part, because they disrupt
stems in RNA secondary structures. Furthermore, we provide evidence that these
mutations are brought to high frequency through linkage to adaptive mutations.

Contrary to naive expectations, the adaptive escape mutations don't seem to be
unconditionally beneficial. Otherwise we should observe almost sure fixation of
non-synonymous mutations once they reach intermediate frequency. Instead, we
find that the fixation probability of non-synonymous mutations is roughly given
by its frequency. There are several possible explanations for this observation. 
Similar to synonymous mutations, the majority of non-synonymous mutations could
be weakly deleterious and the adaptive and deleterious parts conspire to yield
the neutral-like averaged fixation probability. We consider this possibility
inplausible since the amino-acid sequence outside the variable loops is much
more conserved that the synonymous positions, suggesting that the majority of
the non-synonymous mutations is much more deleterious.

Alternatively, the lack of fixation could be due to time dependent environment
through an immune system that is catching up, or competition between mutations
that mediate escape within the same epitope. We explored both of these
possibilities and find that both produce the desired effect. Furthermore, there
is evidence in support of both of these hypothesis. Serum from HIV infected
individuals tyically neutralizes the virus that dominated the population a few
(3-6) month ago \citep{richman_rapid_2003}. This suggests that escape mutations
cease to be beneficial after a few month and might revert if they come with a
fitness cost. Deep sequencing of regions of \env{} after antibody escape have
revealed multiple escape mutations in the same epitope
\citep{moore_limited_2009, bar_early_2012}. Presumably, each one of these
mutations is sufficient for escape but most combinations of them don't provide
any additional benefit to the virus. Hence only one of them will spread and the
others will be driven out of the population even if they transiently reach high
frequencies. The rapid emergence of multiple escape mutations in the same
epitope implies a large population size that explores all necessary point mutations rapidly. A
similar point has been made recently by Boltz {\it et al.} in the context of
preexisting drug resistance mutations~\citep{boltz_ultrasensitive_2012}.

The observed hitchhiking highlights the importance of linkage due to infrequent
recombination for the evolution of HIV
\citep{neher_recombination_2010,batorsky_estimate_2011,
josefsson_majority_2011}. The recombination rate has been estimated to be on the
order of $\rho = 10^{-5}$ per base and day. It takes roughly $t_{sw} =
\epsilon^{-1} \log \nu_0$ generations for escape mutation with escape rate
$\epsilon$ to rise from an initially low frequency $\nu_0\sim \mu$ to frequency
one. This implies that a region of length $l = (\rho t_{sw})^{-1} = \epsilon /
\rho \log \nu_0$ remains linked to the adaptive mutation. With $\epsilon=0.01$,
we have $l\approx 100$ bases. Hence we expect strong linkage between the
variable loops and the intervening sequence, but none far beyond the variable
regions, consistent with the lack of signal outside of C1-V5. In case of much
stronger selection -- such as observed during early CTL escape or drug
resistance evolution -- the linked  region is of course much larger
\citep{nijhuis_stochastic_1998}.

A functional significance of the insulating RNA structure stems between the
hyper-variable loops has been proposed
previously~\citep{watts_architecture_2009, sanjuan_interplay_2011}.
Our analysis is limited by the availability of
longitudinal data which requires a focus on the the variable regions of \env.
Conserved RNA structures exist in different parts of the HIV genome (several are
known). In absence of repeated adaptive substitutions in the vicinity that cause
hitchhiking, the deleterious synonymous mutations remain at low frequencies and
can only be observed by deep sequencing methods.

@@

Our results emphasize the inadequacy of independent site
models of HIV evolution and the common assumption that selection is time
independent or additive. If genetic variation is only transiently beneficial,
existing methods to quantify selection will yield substantial underestimates
\citep{williamson_adaptation_2003,neher_rate_2010,batorsky_estimate_2011}.

\section{Methods}
\subsection{Sequence data collection}
Longitudinal intrapatient viral RNA sequences were collected for published
studies~\citep{shankarappa_consistent_1999,
liu_selection_2006, bunnik_autologous_2008} and downloaded from the Los Alamos
National Laboratory (LANL) HIV sequence database~\citep{LANL2012}. The sequences from some patients
showed signs of HIV compartimentalization into subpopulations and were discarded; a total of 11
patients with approximately 6 time points each and 10 sequences per time point
were analyzed. The time interval or resolution between two consecutive sequences
was approximately 6 to 18 months.

@@

\subsection{Sequence analysis}
The good sequences were aligned within each patient
via the translated amino acid sequence, using
Muscle~\citep{edgar_muscle:_2004}, and to the NL4-3 reference sequence used
by \citet{watts_architecture_2009} in the SHAPE assay. Within each patient, a
consensus nucleotide sequence at the first time point was used to classify
alleles as ancestral or derived at all sites. Problematic sites that included large frequencies of gaps
were excluded from the analysis to avoid artefactual substitutions due to
alignment errors. Time series of allele frequencies were extracted from the
sequences. \comment{The alignments used are available for download}.

@@

\comment{All analysis scripts are available for download\ldots.}. 
\section*{Acknowledgements}
We are grateful for stimulating discussions with Jan Albert and Trevor Bedford.
This work is supported by the ERC starting grant HIVEVO 260686.
